A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-06-08 paper/paper.tex
Experimental evaluation section.

In this paper, we propose a new heuristic algorithm for the crossing resolution maximization problem and we experimentally compare it against the known approaches from the literature. Our experimental evaluation indicates that the new heuristic produces drawings with significantly better crossing resolution, but this comes at the cost of slightly higher running time, especially when the input graph is large and dense. 
\end{abstract}

@@

\myparagraph{Crossing resolution} Our results for the crossing resolution are summarized in Fig.~\ref{fig:cr-res}. It is immediate to see that our algorithm outperforms all other ones in terms of the crossing resolution of the produced drawings, when we do not impose any restriction on the aspect ratio of the computed drawings; refer to the solid-black curve, denoted as \emph{unrestricted}, in Fig.~\ref{fig:cr-res-1}. The variant of our algorithm, which does not violate the aspect ratio of the initial layout, leads to drawings with slightly smaller crossing resolution; refer to the solid-gray curve, denoted as \emph{ar-restricted}, in Fig.~\ref{fig:cr-res-1}. Finally, the two spring-embedder algorithms seem to produce drawings with worse crossing resolution; refer to the dotted-black and dotted-gray curves of Fig.~\ref{fig:cr-res-1}.

While our unrestricted algorithm produces drawings of better crossing resolution, this seems to come at a cost of increased aspect ratio (see Fig.~\ref{fig:cr-res-2}), which, however, is still better that the corresponding aspect ratio of the drawings produced by the algorithm of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}. There it seems that the forces due to angles formed at the crossings outperform the corresponding spring forces, which try to keep the lengths of the edges short. Regarding our unrestricted algorithm, its behavior is up to a certain degree expected mainly due to the fact that there is no control on the lengths of the edges (as it is, e.g., in the case of the force-directed algorithm of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}, whose sping forces inpose stong restrictions on the edge lengths). On the other hand, the restricted variant of our algorithm, which does not allow the aspect ratio to decrease, has more or less comparable performace (in terms of the aspect ratio of the computed drawings) as the one of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}.

Regarding the number of crossings of the computed drawings, we observe that the restricted variant of our algorithm and the spring embedder algorithm of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} yield drawings with comparable number of crossings, which at the same time is significantly smaller than the corresponding number of crossings produced by the two other algorithms of our expertiment; refer to Fig.~\ref{fig:cr-res-3}.

A different behavior can be observed in the number of iterations, which are required by the algorithms of our experiments, in order to converge; refer to Fig.~\ref{fig:cr-res-4}. Note here that we used different criteria to determine whether our algorithm (and its variant) and whether the two spring embedder algorithms have converged, mainly due to the different nature of these algorithms. For each of the two spring embedder algorithms, if the crossing resolution between two consecutive iterations was not improved more than 0.001 degrees, we assumed that the algorithm had converged and we did not proceed any more. On the other hand, for both of our algorithms we used a more relaxed criterion, because these algorithms may perform several iterations without changing the drawing (and thus, its crossing resolution), as they based on randomization. However, if after 100 consecutive iterations there was no improvement, then we assumed that the algorithm had converged and we did not proceed any more. The maximum number of iterations that each of the algorithms of our experiment could perform in order to converge was set to 100.000. By Fig.~\ref{fig:cr-res-3}, it is immediate to see that both the unrestricted and the restricted variant of our algorithm require less iterations than the two spring-embedder algorithms.

%. If both the angular and the
%crossing resolution between two consecutive iterations of each
%algorithm were not improved more than 0.001 degrees, we assumed that
%the algorithm has converged and we did not proceed any more. We note
%that the termination condition is quite strict and demands a large
%number of iterations. 
%However, very few graphs required the maximum number of iterations
%in order to converge to a layout. Once an algorithm converged, we
%measured its angular, crossing and total resolution, its average
%edge length, the number of crossings, the running time, and the
%required number of iterations to converge.

\begin{itemize}
\item Results.
\item Discussion.
\end{itemize}

>>>>

In this paper, we propose a new heuristic algorithm for the crossing resolution maximization problem and we experimentally compare it against the known approaches from the literature. Our experimental evaluation indicates that the new heuristic produces drawings with better crossing resolution, but this comes at the cost of slightly higher aspect ratio, especially when the input graph is large. 
\end{abstract}

@@

\myparagraph{Crossing resolution} Our results for the crossing resolution are summarized in Fig.~\ref{fig:cr-res}. Here, each algorithm was adjusted so to maximize exclusively the crossing resolution (i.e., by ignoring the drawing's angular resolution). It is immediate to see that our algorithm outperforms all other ones in terms of the crossing resolution of the produced drawings, when we do not impose any restriction on the aspect ratio of the computed drawings; refer to the solid-black curve, denoted as \emph{unrestricted}, in Fig.~\ref{fig:cr-res-1}. The variant of our algorithm, which does not violate the aspect ratio of the initial layout, leads to drawings with slightly smaller crossing resolution; refer to the solid-gray curve, denoted as \emph{ar-restricted}, in Fig.~\ref{fig:cr-res-1}. Finally, the two force-directed algorithms seem to produce drawings with worse crossing resolution; refer to the dotted-black and dotted-gray curves of Fig.~\ref{fig:cr-res-1}.

While our unrestricted algorithm produces drawings of better crossing resolution, this seems to come at a cost of increased aspect ratio (see Fig.~\ref{fig:cr-res-2}), which, however, is still better that the corresponding aspect ratio of the drawings produced by the algorithm of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}. For the latter algorithm, it seems that the forces due to angles formed at the crossings outperform the corresponding spring forces, which try to keep the lengths of the edges short. Going back to our unrestricted algorithm, its behaviour is up to a certain degree expected mainly due to the fact that there is no control on the lengths of the edges.
% (as it is, e.g., in the case of the force-directed algorithm of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}, whose sping forces inpose stong restrictions on the edge lengths). 
On the other hand, the restricted variant of our algorithm, which does not allow the aspect ratio to decrease, has more or less comparable performace (in terms of the aspect ratio of the computed drawings) as the one of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}.

Regarding the number of crossings, we observe that the restricted variant of our algorithm and the force-directed algorithm of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} yield drawings with comparable number of crossings, which at the same time is significantly smaller than the corresponding number of crossings produced by the two other algorithms of our experiment; refer to Fig.~\ref{fig:cr-res-3}.

A different behaviour can be observed in the number of iterations, which are required by the algorithms to converge; refer to Fig.~\ref{fig:cr-res-4}. We note here that we used different criteria to determine whether our algorithm (and its variant) and whether the two force-directed algorithms have converged, mainly due to the different nature of these algorithms. For each of the two force-directed algorithms, if the crossing resolution between two consecutive iterations was not improved more than 0.001 degrees, we assumed that the algorithm had converged and we did not proceed any more. On the other hand, for both of our algorithms we used a more relaxed criterion, because these algorithms may perform several iterations without changing the drawing (and thus, its crossing resolution), as they based on randomization. However, if after 100 consecutive iterations there was no improvement, then we assumed that the algorithm had converged and we did not proceed any more. The maximum number of iterations that each of the algorithms of our experiment could perform in order to converge was set to 100.000. By Fig.~\ref{fig:cr-res-4}, it is immediate to see that both the unrestricted and the restricted variant of our algorithm require less iterations than the two force-directed algorithms.

\myparagraph{Total resolution} Our results for the total resolution are summarized in Fig.~\ref{fig:to-res}. Here, each algorithm was adjusted so to maximize the minimum of the crossing and of the angular resolution. We observe that both variants of our algorithm as well as the force-directed algorithm by Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} tend to produce drawings of the same total resolution. Slightly worse is the performance of the algorithm by Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}; see Fig.~\ref{fig:to-res-1}. However, the aspect ratio of the drawings of the latter algorithm is better than the corresponding aspect ratios of the remaining algorithms of the experiment; see Fig.~\ref{fig:to-res-2}. More concritely, the drawings produced by the restricted variant of our algorithm have slightly worse aspect ratios. Then, the ones produced by the force directed algorithm of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} follow. However, we again observe that our unrestricted algorithm leads to drawings with very high aspect ratio.

Similar observations can be made also for the number of crossings of the produced layouts; see Fig.~\ref{fig:to-res-3}. The algorithm by Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} yields drawings with the least number of crossings. Comparable but slightly worse (in terms of the number of crossings) are the drawings produced by the restricted variant of our algorithm and by the force-directed algorithm by Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}. Our unrestricted algorithm seems to require the largest number of crossings, which turns out to be notably higher than the corresponding ones of the remaining algorithms of our experiment especially for large graphs. On the positive side, both the unrestricted and the restricted variant of our algorithm require less iterations than the two force-directed algorithms, as it can be easily seen in Fig.~\ref{fig:to-res-4}.

\myparagraph{Angular resolution} We conclude the analysis of our experimental evaluation with the  results for the angular resolution; refer to Fig.~\ref{fig:an-res}. Here, each algorithm was adjusted  to maximize exclusively the angular resolution (i.e., by ignoring the drawing's crossing resolution). A notable observation is that the force-directed algorithm by Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} tends to have the same performance (in term of angular resolution) as our unrestricted algorithm, which is slightly better but mostly for small graphs; see Fig.~\ref{fig:an-res-1}. The restricted variant of our algorithm yields drawings with slightly worse angular resolution. The force-directed algorithm by Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} seems to be outperformed by all algorithms of the experiment.

On the other hand, the results for the aspect ratio, the number of crossings and the required number of iterations are very similar with corresponding ones for the total resolution; see Figs.~\ref{fig:an-res-2}--\ref{fig:an-res-4}. This observation suggests that, for most of the graphs of our experiment, the angular resolution dominates the crossing resolution (and thus is the one defining the total resolution) in the constructed drawings, which explains the similarity in the reported results.
